---
title: "Coursera - Machine Learning Final Assessment"
author: "Adrian Marcia"
date: "February 12, 2018"
output: html_document
---

#Machine Learning Final Assessment

##Overview

###Project background
Personal activity data is increasing in availability. While the quantity of data is increasing, a measure of quality is not readily available.
Using data from 6 individuals, this project aims to determine which factors align to a successful barbell lift.

###Data
The training data was obtained from: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
The test data was obtained from: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

###Objective
Determine which factors impact whether a barbell lift is performed correctly.

### Library packages
Installing and loading the packages for machine learning and plotting
```{r}
library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
library(RColorBrewer)
```

### Data retrieval
Downloading and storing the data
```{r}
#Set seed at 1 to create reproducibility
set.seed(1)

trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

training_data <- read.csv(url(trainUrl), na.strings=c("NA","#DIV/0!",""))
testing_data <- read.csv(url(testUrl), na.strings=c("NA","#DIV/0!",""))
```

### Data cleaning
```{r}
##Remove nearzerovariance variables
nzv <- nearZeroVar(training_data, saveMetrics=TRUE)
training_data <- training_data[,nzv$nzv==FALSE]

##Remove the first column
training_data <- training_data[c(-1)]

##Remove variables with full of NA and make testing dataset consistent with training dataset in terms of available columns
features <- names(training_data[,colSums(is.na(training_data)) == 0])[1:57]
training_data<-training_data[,c(features,"classe")]
testing_data<-testing_data[,c(features)]

for (i in 1:length(testing_data) ) {
  for(j in 1:length(training_data)) {
    if( length( grep(names(training_data[i]), names(testing_data)[j]) ) ==1)  {
      class(testing_data[j]) <- class(training_data[i])
    }      
  }      
}

testing_data <- rbind(training_data[2, -58] , testing_data) 
testing_data <- testing_data[-1,]
```

##Data partitioning
```{r}
inTrain <- createDataPartition(y=training_data$classe, p=0.6, list=FALSE)
mytraining <- training_data[inTrain, ]; mytesting <- training_data[-inTrain, ]
dim(mytraining)
```

##Prediction with Decision Trees
```{r}
DT <- rpart(classe ~ ., data=mytraining, method="class")
rpart.plot(DT)
```

##Confusion matrix
```{r}
DT2 <- predict(DT, mytesting, type = "class")

#additional packages required for the confusion matrix
#install.packages('e1071', dependencies=TRUE)
library("e1071")

confusionMatrix(DT2, mytesting$classe)
```

##Random forest prediction
```{r}
RF <- randomForest(classe ~. , data=mytraining)
RF2 <- predict(RF, mytesting, type = "class")
confusionMatrix(RF2, mytesting$classe)
```

##Out of sample error
```{r}
predictions <- predict(RF, testing_data, type = "class")
predictions
```
